A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-02-07 sections/algorithm.tex
Section 3

This section introduces a novel algorithmic system that implements the
DK's declarative specification. The algorithmic system comes with
several novelties.\bruno{fill in/summarize novelties}

@@

The design of our judgment chain is closely related to the property of
the judgments\bruno{What property? Not very clear}.
Subtyping and checking do not return anything, so variables cannot
leak anyway, as applied to Rules 8 and 9.
\bruno{Is the discussion that follows in the right place? We just to
  rule 26. Perhaps we can wait until we talk about inference to
  discuss those issues?}
However, inference and application inference may return a type that contains new variables.
Take Rule 26 as an example, if it simply returns $\al \to \bt$, passes that to the next judgment and continues the type-checking process,
variables $\al$ and $\bt$ in the next judgment are out of scope
and will break the well-formedness of the worklist.
By making use of the continuation-passing-style judgement chain,
inner judgments always shares context with its parent judgment,
therefore we preserve well-formedness during reduction.

Secondly, Rules 10 and 11 are algorithmic extensions to Rule $\mathtt{{\le}{\to}}$.
Both of them partially instantiate $\al$ to function types.
The domain $\al[1]$ and range $\al[2]$ of the new function type are not determined:
they are fresh existential variables with the same scope as $\al$.
We remove $\al$ from the worklist, and insert $\al[1], \al[2]$ back to the same position.
To propagate the instantiation to the rest of the worklist and maintain the wel-formedness,
every $\al$ is replaced by $\al[1] \to \al[2]$.
The \emph{occurs-check} condition is to prevent a diverging infinite loop.
Without such check, $\al \le 1 \to \al$, as an example, would diverge.

Thirdly, Rules 12-17 are base cases where an existential variable is solved.
Each of the rules removes and existential variable, substitutes the variable with its solution and continues.

Fourthly, Rules 18-21 deal with checking judgments.
Rule 18 is similar to $\mathtt{DeclSub}$, and rewrited in the continuation-passing-style.
Rules 19 and 20 adapts the worklist style from their corresponding declarative rules.
Rule 21 is a special case of $\mathtt{Decl\to I}$.
Instantiation technique similar to Rules 10 and 11 applys.
The declarative checking rule $\mathtt{Decl1I}$ is not translated to the algorithm,
since combining Rule 18 and Rule 24 would give the same result.

Fifthly, Rules 22-26 deal with inference judgments.
Inference judgment chains are written in continuation-passing-style.
When an inference judgment succeeds with type $A$,
the algorithm continues to work on the inner-chain $\jg$ by assigning $A$ to its placeholder varible $a$.
Rule 23 infers an annotated expression by changing into checking mode,
therefore another judgment chain is created.
Rule 25 infers the type of a lambda expression by introducing $\al, \bt$
as the input and output types of the function, respectively.
After checking the body $e$ under the assumption $x:\al$,
the return type might reflect more information than simply $\al \to \bt$.
The variable scopings are well-maintained during the process:
the assumption of argument type ($x:\al$) is recycled after checking against the function body;
the declaration of the function type ($\al,\bt$) is only visible to
the checking and the remaining chain $[\al\to\bt/a]\jg$.

Finally, Rules 27-29 deal with application inference judgments.
Rules 27 and 28 behaves similarly to declarative rules $\mathtt{Decl\forall App}$ and $\mathtt{Decl\to App}$.
Rule 29 instantiate $\al$ to the function type $\al[1] \to \al[2]$, just like Rules 10, 11 and 21.

The syntax $\Gm\redto \Gm'$ denotes multiple reduction.
If $\Gm\redto\nil$, we say that $\Gm$ reduces successfully.

@@

Figure~\ref{fig:alg:sample} shows a sample derivation of the algorithm.

>>>>

This section introduces a novel algorithmic system that implements the
DK's declarative specification. The algorithmic system comes with
several novelties.\bruno{fill in/summarize novelties. I think
  novelties include: precise scoping control; CPS-style; dealing with
  inference judgements.}

@@

The design of our judgment chain is closely related to the property of
the judgments\bruno{What property? Not very clear}.
Subtyping and checking do not return anything, so variables cannot
leak anyway, as applied to Rules 8 and 9.
\bruno{Is the discussion that follows in the right place? We just jump
  to
  rule 26. Perhaps we can wait until we talk about inference to
  discuss those issues?}
However, inference and application inference may return a type that contains new variables.
Take Rule 26 as an example, if it simply returns $\al \to \bt$, passes that to the next judgment and continues the type-checking process,
variables $\al$ and $\bt$ in the next judgment are out of scope
and will break the well-formedness of the worklist.
By making use of the continuation-passing-style judgement chain,
inner judgments always shares context with its parent judgment,
therefore we preserve well-formedness during reduction.

\paragraph{Existential decomposition}
Secondly, Rules 10 and 11 are algorithmic extensions to Rule $\mathtt{{\le}{\to}}$.
Both of them partially instantiate $\al$ to function types.
The domain $\al[1]$ and range $\al[2]$ of the new function type are not determined:
they are fresh existential variables with the same scope as $\al$.
We remove $\al$ from the worklist, and insert $\al[1], \al[2]$ back to the same position.
To propagate the instantiation to the rest of the worklist and maintain the well-formedness,
every $\al$ is replaced by $\al[1] \to \al[2]$.
The \emph{occurs-check} condition is to prevent a diverging infinite loop.
Without such check, $\al \le 1 \to \al$, as an example, would diverge.

\paragraph{Solving existentials.} Thirdly, Rules 12-17 are base cases where an existential variable is solved.
Each of the rules removes and existential variable, substitutes the
variable with its solution and continues. Importantly the rules
respect the scoping of existential variables. For example, rule 12 
states that an existential variable $\al$ can be solved with another
existential variable $\bt$ only if $\bt$ occurs after $\al$.

\paragraph{Checking judgements}
Fourthly, Rules 18-21 deal with checking judgments.
Rule 18 is similar to $\mathtt{DeclSub}$, and rewrited in the
continuation-passing-style.
\bruno{The side condition needs to be explained, and some mention of
  the overlapping is probably needed here.}
Rules 19 and 20 adapts the worklist style from their corresponding declarative rules.
Rule 21 is a special case of $\mathtt{Decl\to I}$.\bruno{Rule 21
  deserves some more explanation. What is the motivation for this
  rule? You may want to give some concrete example to motivate it.}
Instantiation technique similar to Rules 10 and 11 applies.
The declarative checking rule $\mathtt{Decl1I}$ is not translated to the algorithm,
since combining Rule 18 and Rule 24 would give the same result.

\paragraph{Inference judgments}
Fifthly, Rules 22-26 deal with inference judgments.
Inference judgment chains are written in continuation-passing-style.
When an inference judgment succeeds with type $A$,
the algorithm continues to work on the inner-chain $\jg$ by assigning $A$ to its placeholder varible $a$.
Rule 23 infers an annotated expression by changing into checking mode,
therefore another judgment chain is created.
Rule 25 infers the type of a lambda expression by introducing $\al, \bt$
as the input and output types of the function, respectively.
After checking the body $e$ under the assumption $x:\al$,
the return type might reflect more information than simply $\al \to \bt$.
The variable scopings are well-maintained during the process:
the assumption of argument type ($x:\al$) is recycled after checking against the function body;
the declaration of the function type ($\al,\bt$) is only visible to
the checking and the remaining chain $[\al\to\bt/a]\jg$.
\bruno{You have some note ``differs'' on the rules. Should that
  deserve an explanation here?}

\paragraph{Application inferences judgements}
Finally, Rules 27-29 deal with application inference judgments.
Rules 27 and 28 behaves similarly to declarative rules $\mathtt{Decl\forall App}$ and $\mathtt{Decl\to App}$.
Rule 29 instantiate $\al$ to the function type $\al[1] \to \al[2]$, just like Rules 10, 11 and 21.

\subsection{Multiple reduction}
The syntax $\Gm\redto \Gm'$ denotes multiple reduction.
If $\Gm\redto\nil$, we say that $\Gm$ reduces successfully.

@@

Figure~\ref{fig:alg:sample} shows a sample derivation of the
algorithm.
\bruno{Derivation should be explained briefly at least.}
